---
title: "Programming languages history"
author: "Flavio Blondeau"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### Data on tags per year

We want to investigate what are the most used programming languages over the years. To do so, we look at questions and answers posted on <https://stackoverflow.com/>. We will use a dataset containing observations for each **tag** in each year.

### First look at data

We load the tidyverse library

```{r}
#| label: tidyverse
#| message: false
library(tidyverse)
```

and then we load the dataset and give a first look to the data

```{r}
#| label: dataset
#| message: false
by_tag_year <- read_csv("by_tag_year.csv")
by_tag_year
```

Along with **year** and **tag**, we have two other variables, **number** and **year_total**. The first one counts the number of questions posted with the tag, and the second one is the total number of questions in the year.

### Tag percentage

We want to look at the percentage of tag questions over total ones (per year)

```{r}
#| label: percentage

by_tag_year_perc <- by_tag_year |> 
  mutate(
    percentage = round(number /year_total * 100, 2)
  )
by_tag_year_perc
```

What are the most used tags for each year?

```{r}
#| label: most-used-tag

by_tag_year_perc |> 
  group_by(year) |> 
  arrange(desc(percentage)) |> 
  filter(row_number() == 1) |> 
  arrange(year)
```

We can see that the most used tags are *c#* from 2008 to 2011 and *javascript* from 2013 to 2018, while 2012 winner was *java*.


### Most used programming languages

We can expand our analysis by considering the top 3 languages for each year
```{r}
#| label: most-used-3-tags

top_three <- by_tag_year_perc |> 
  group_by(year) |> 
  arrange(desc(percentage)) |> 
  filter(row_number() %in% c(1,2,3)) |> 
  arrange(year)
top_three
```

Since the table is too long to be seen entirely, we plot a bar graph

```{r}
#| label: top-three-plot

top_three |> 
  ggplot(aes(x = year, y = percentage/100, fill = tag)) +
  geom_bar(position="dodge", stat="identity") +
  scale_x_continuous(breaks = 2008:2018) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Top 3 programming languages between 2008 and 2018", x = "Year", y = "R tags (percentage)", color = "Tags") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```



### What about R?

Let's see the percentages about R

```{r}
#| label: R-data

r_over_time <- by_tag_year_perc |> 
  filter(tag == "r")
r_over_time
```

We can plot a graph to better understand the trend

```{r}
#| label: R-plot

r_over_time |> 
  ggplot(aes(x = year, y = percentage/100)) +
  geom_point() +
  geom_line() +
  scale_x_continuous("Year", breaks = 2008:2018) +
  scale_y_continuous("R tags (percentage)", labels = scales::percent)
```

It seems that R language has become more popular over the years.
